\thornado-hydro: towards discontinuous galerkin methods for supernova hydrodynamics

### 2018-12-14 Documents/Euler/Astronum_2018/ms_endeve.tex
Progress on reviewer comments

\newcommand{\deriv}[2]{\frac{d #1}{d #2}}
\newcommand{\pderiv}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\vect}[1]{\boldsymbol{#1}}
\newcommand{\f}[2]{\frac{#1}{#2}}
\newcommand{\dx}{\Delta x}
\newcommand{\pd}[2]{\partial_{#2}{#1}}
\newcommand{\bx}{\vect{x}}
\newcommand{\bK}{\vect{K}}
\newcommand{\cT}{\mathcal{T}}
\newcommand{\xL}{x_{\textnormal{\tiny\textsc{L}}}}
\newcommand{\xH}{x_{\textnormal{\tiny\textsc{H}}}}
\newcommand{\xMin}{x_{\textnormal{\tiny m}}}
\newcommand{\xMax}{x_{\textnormal{\tiny M}}}
\newcommand{\trans}{\textnormal{\tiny\textsc{T}}}
\newcommand{\leftState}{\textnormal{\tiny\textsc{L}}}
\newcommand{\rightState}{\textnormal{\tiny\textsc{R}}}
\newcommand{\TCI}{\textnormal{\tiny\textsc{TCI}}}
\newcommand{\shock}{\textnormal{\tiny\textsc{Sh}}}
\newcommand{\thornado}{\texttt{thornado}}
\newcommand{\chimera}{{\sc Chimera}}

@@

@@

Here we show results from solving the SR Euler equations in Cartesian coordinates with \thornado.
In this case the state, flux, and source vectors in Eq.~\eqref{eq:extendedEulerCompact} are
\begin{equation}
  \vect{U}=\big(\rho W, \rho h W^{2} u_{j},\tau\big)^{\trans},~
  \vect{F}^{i}=\big(\rho W u^{i},\Pi^{i}_{~j},\rho(hW-1)W u^{i}\big)^{\trans},~
  \text{and}~\vect{S}=0,
\end{equation}
where $W$ is the Lorentz factor, $h=1+(e+p)/\rho$ is the specific enthalpy, $\tau=\rho W(hW-1)-p$, and $\Pi^{i}_{~j}=\rho h W^{2} u^{i} u_{j}+p\delta^{i}_{~j}$.  
We solve two Riemann problems and compare the numerical results with exact solutions obtained with the solver from \cite{martiMuller_2003}.  
Let the primitive state vector be $\vect{V}=\big(\rho,\vect{u},p\big)^{\trans}$.  
In the first problem (Riemann problem~1), taken from \cite{mignoneBodo_2005}, we use $\Gamma=4/3$ and
\begin{equation*}
  \vect{V}_{\leftState}=\big(1.0,0.9,0.0,0.0,1.0\big)^{\trans}\quad\text{and}\quad
  \vect{V}_{\rightState}=\big(1.0,0.0,0.0,0.0,10.0\big)^{\trans},
\end{equation*}
while in the second problem (Riemann problem~2), also from \cite{mignoneBodo_2005}, we use $\Gamma=5/3$, and let
\begin{equation*}
  \vect{V}_{\leftState}=\big(1.0,\vect{0.0},10^{3}\big)^{\trans}\quad\text{and}\quad
  \vect{V}_{\rightState}=\big(1.0,\vect{0.0},10^{-2}\big)^{\trans}.  
\end{equation*}
In both problems, the computational domain is $D=[0,1]$, the discontinuity is initially located at $x^{1}=0.5$, and the solutions are integrated to $t=0.4$.  
Results are plotted in Figure~\ref{fig:RelativisticHydro}.

\begin{figure}[h]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/MB2005_01_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/MB2005_04_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:RelativisticHydro}Results from solving the SR Euler equations with \thornado.  Riemann problem~1 (left panel) was solved using $100$ elements, while Riemann problem~2 was solved using $400$ elements.  The numerical and exact solutions are plotted with plusses and solid lines, respectively.}
\end{figure}

Again, the main features of the exact solution are captured with the DG method implemented in \thornado.  
For Riemann problem~1, we observe some oscillations in the density profile between the leftmost shock and the contact discontinuity located around $x^{1}=0.6$.  
For Riemann problem~2, the maximum density in the thin density shell between the shock and the contact discontinuity is significantly lower for the numerical solution than the exact solution.  
This is due to a combination of limited spatial resolution and the action of the slope limiter.  
We used $C_{\TCI}=0.03$ in both tests in this section.  
We have found the agreement with the exact solution to improve for larger values of $C_{\TCI}$, but at the expense of somewhat more oscillatory results.

@@

\section*{References}
\begin{thebibliography}{36}
  \bibitem{ascher_etal_1997} Ascher U M, Ruuth S J and Spiteri R J 1997 {\it Applied Numerical Mathematics} {\bf 23} 151
  \bibitem{bassi_etal_2013} Bassi F, Franchina N, Ghidoni A and Rebay S 2013 {\it Int. J. Numer. Meth. Fluids} {\bf 71} 1322
  \bibitem{blondin_etal_2003} Blondin J M, Mezzacappa A and DeMarino C 2003 {\it ApJ} {\bf 584} 980
  \bibitem{brennerScott_2008} Brenner S C and Scott L R 2008 {\it The Mathematical Theory of Finite Element Methods} Springer
  \bibitem{bruenn_etal_2018} Bruenn S W {\it et al} {\it Preprint} arXiv:1809.05608
  \bibitem{cockburnShu_1998} Cockburn B and Shu C-W 1998 {\it JCP} {\bf 141} 199
  \bibitem{cockburnShu_2001} Cockburn B and Shu C-W 2001 {\it Journal of Scientific Computing} {\bf 16} 173
  \bibitem{colellaWoodward_1984} Colella P and Woodward P 1984 {\it JCP} {\bf 54} 174
  \bibitem{dumbser_etal_2014} Dumbser M, Zanotti O, Loub{\`e}re R and Diot S 2014 {\it JCP} {\bf 278} 47
  \bibitem{fambri_etal_2018} Fambri F, Dumbser M, K{\"o}ppel S, Rezzolla L and Zanotti O 2018 {\it MNRAS} {\bf 477} 4543
  \bibitem{fuShu_2017} Fu G and Shu C-W 2017 {\it JCP} {\bf 347} 305
  \bibitem{harten_etal_1983} Harten A, Lax P D and Van Leer B 1983 {\it SIAM~Review} {\bf 25} 35
  \bibitem{hesthavenWarburton_2008} Hesthaven J S and Warburton T 2008 {\it Nodal discontinuous Galerkin methods: Algorithms, analysis and applications} Springer
  \bibitem{klockner_etal_2009} Kl{\"o}ckner A, Warburton T, Bridge J and Hesthaven J S 2009 {\it JCP} {\bf 228} 7863
  \bibitem{landauLifshitz_1979} Landau L D and Lifshitz E M 1979 {\it Fluid Mechanics} Pergamon
  \bibitem{liskaWendroff_2003} Liska R and Wendroff B 2003 {\it SIAM J. Sci. Comput.} {\bf 25} 3
  \bibitem{martiMuller_2003} Mart{\'i} J M and M{\"u}ller E 2003 {\it Living Rev. Relativ.} {\bf 6} 7
  \bibitem{mcnally_etal_2012} McNally C, Lyra W and Passy J-C 2012 {\it ApJS} {\bf 201} 18
  \bibitem{mignoneBodo_2005} Mignone A and Bodo G 2005 {\it MNRAS} {\bf 364} 126
  \bibitem{muller_2016} M{\"u}ller B 2016 {\it PASA} {\bf 33} 1
  \bibitem{qin_etal_2016} Qin T, Shu C-W and Yang Y 2016 {\it JCP} {\bf 315} 323
  \bibitem{radiceRezzolla_2011} Radice D and Rezzolla L 2011 {\it Phys. Rev. D} {\bf 84} 024010
  \bibitem{remacle_etal_2003} Remacle J-F, Flaherty J E and Shephard M S 2003 {\it SIAM~Review} {\bf 45} 53
  \bibitem{schaal_etal_2015} Schaal K, Bauer A, Chandrashekar P R{\"u}diger P, Klingenberg C and Springel V 2015 {\it MNRAS} {\bf 453} 4278
  \bibitem{sedov_1959} Sedov L I 1959 {\it Similarity and Dimensional Methods in Mechanics} Academic Press
  \bibitem{shuOsher_1988} Shu C-W and Osher S 1988 {\it JCP} {\bf 77} 439
  \bibitem{shuOsher_1989} Shu C-W and Osher S 1989 {\it JCP} {\bf 83} 32
  \bibitem{shu_1997} Shu C-W 1997 {\it NASA/CR-97-206253 ICASE Report No. 97-65}
  \bibitem{shu_2016} Shu C-W 2016 {\it JCP} {\bf 316} 598
  \bibitem{sod_1978} Sod G A 1978 {\it JCP} {\bf 27} 1
  \bibitem{teukolsky_2016} Teukolsky S A 2016 {\it JCP} {\bf 312} 333
  \bibitem{toro_etal_1994} Toro E F, Spruce M and Speares W 1994 {\it Shock Waves} {\bf 4} 25
  \bibitem{toro_1999} Toro E F 1999 {\it NUMERICA: A Library of Source Codes for Teaching, Research and Applications}
  \bibitem{wilson_etal_1996} Wilson J R, Mathews G J and Marronetti P 1996 {\it Phys. Rev. D} {\bf 54} 1317
  \bibitem{wuTang_2015} Wu K and Tang H 2015 {\it JCP} {\bf 298} 539
  \bibitem{yahilLattimer_1982} Yahil A and Lattimer J M 1982 {\it Supernovae: A Survey of Current Research} Reidel
  \bibitem{yahil_1983} Yahil A 1983 {\it ApJ} {\bf 265} 1047
  \bibitem{zhangShu_2010} Zhang X and Shu C-W 2010 {\it JCP} {\bf 229} 8918
\end{thebibliography}

>>>>

\newcommand{\deriv}[2]{\frac{d #1}{d #2}}
\newcommand{\pderiv}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\vect}[1]{\boldsymbol{#1}}
\newcommand{\f}[2]{\frac{#1}{#2}}
\newcommand{\dx}{\Delta x}
\newcommand{\pd}[2]{\partial_{#2}{#1}}
\newcommand{\bx}{\vect{x}}
\newcommand{\bK}{\vect{K}}
\newcommand{\cT}{\mathcal{T}}
\newcommand{\xL}{x_{\textnormal{\tiny\textsc{L}}}}
\newcommand{\xH}{x_{\textnormal{\tiny\textsc{H}}}}
\newcommand{\xMin}{x_{\textnormal{\tiny m}}}
\newcommand{\xMax}{x_{\textnormal{\tiny M}}}
\newcommand{\trans}{\textnormal{\tiny\textsc{T}}}
\newcommand{\leftState}{\textnormal{\tiny\textsc{L}}}
\newcommand{\rightState}{\textnormal{\tiny\textsc{R}}}
\newcommand{\TCI}{\textnormal{\tiny\textsc{TCI}}}
\newcommand{\shock}{\textnormal{\tiny\textsc{Sh}}}
\newcommand{\Pos}{\textnormal{\tiny\textsc{Pos}}}
\newcommand{\TVD}{\textnormal{\tiny\textsc{Tvd}}}
\newcommand{\thornado}{\texttt{thornado}}
\newcommand{\chimera}{{\sc Chimera}}

@@

To summarize the RK-DG algorithm we let $\bar{\vect{U}}$ denote the global solution vector containing all the unknowns in $D$ and $\bar{\vect{F}}$ denote the corresponding DG discretization from the right-hand side of Eq.~\eqref{eq:semidiscreteDiscretized}, so that the global ODE system can be written as the $d_{t}\bar{\vect{U}}=\bar{\vect{F}}(\bar{\vect{U}})$.  
We denote the TVD limiter with $\Lambda^{\TVD}\{\}$ and the `positivty' limiter preventing unphysical states with $ \Lambda^{\Pos}\{\}$.  
(E.g., application of the TVD limiter is $\bar{\vect{U}} := \Lambda^{\TVD}\{\bar{\vect{U}}\}$.)  
The general $s$-stage Runge-Kutta time stepping algorithm, which includes application of limiters, can then be written as \cite{cockburnShu_2001}:
\begin{itemize}
  \item[1.] Set $\bar{\vect{U}}^{(0)} = \bar{\vect{U}}^{n}$,
  \item[2.] For $i=1,\ldots,s$ compute:
  \begin{equation}
    \bar{\vect{U}}^{(i)} 
    = \Lambda^{\Pos}\Big\{\,\Lambda^{\TVD}\Big\{\,\sum_{j=0}^{i-1}\Big(\,\alpha_{ij}\,\bar{\vect{U}}^{(j)}+\beta_{ij}\,\Delta t\,\bar{\vect{F}}\big(\bar{\vect{U}}^{(j)}\big)\,\Big)\,\Big\}\,\Big\},
    \label{eq:rkStages}
  \end{equation}
  \item[3.] Set $\bar{\vect{U}}^{n+1}=\bar{\vect{U}}^{(s)}$.  
\end{itemize}
The coefficients $\alpha_{ij}$ and $\beta_{ij}$ for SSP-RK3 are given in Table~2.1 in \cite{cockburnShu_2001}.  
In Eq.~\eqref{eq:rkStages}, for each stage, the TVD limiter is applied first (for elements where $I_{\bK}>C_{\TCI}$), followed by the positivity limiter.  
The positivity limiter

$\Delta t=\min_{D}(\Delta t_{\TVD},\Delta t_{\Pos})$
\begin{equation}
  \Delta t_{\TVD} = \f{1}{d}\,\f{1}{(2k+1)}\min_{i}\big(\Delta x^{i}/|\lambda^{i}|\big)\quad\text{and}\quad
  \Delta t_{\Pos} = \f{\hat{w}_{N}}{d}\min_{i}\big(\Delta x^{i}/|\lambda^{i}|\big)
\end{equation}
For $d=k=2$, the effective Courant-Friedrichs-Lewy (CFL) factor is $0.1$ for $\Delta t_{\TVD}$, while it is $1/12$ for $\Delta t_{\Pos}$.

@@

Here we show results from solving the SR Euler equations in Cartesian coordinates with \thornado.
In this case the state, flux, and source vectors in Eq.~\eqref{eq:extendedEulerCompact} are
\begin{equation}
  \vect{U}=\big(\rho W, \rho h W^{2} u_{j},\tau\big)^{\trans},~
  \vect{F}^{i}=\big(\rho W u^{i},\Pi^{i}_{~j},\rho(hW-1)W u^{i}\big)^{\trans},~
  \text{and}~\vect{S}=0,
\end{equation}
where $W$ is the Lorentz factor, $h=1+(e+p)/\rho$ is the specific enthalpy, $\tau=\rho W(hW-1)-p$, and $\Pi^{i}_{~j}=\rho h W^{2} u^{i} u_{j}+p\delta^{i}_{~j}$.  
To recover primitive from conserved variables we have implemented the procedure for analytic equations of state detailed in Appendix~D in \cite{rezzollaZanotti_2013}.  
We solve two Riemann problems and compare the numerical results with exact solutions obtained with the solver from \cite{martiMuller_2003}.  
Let the primitive state vector be $\vect{V}=\big(\rho,\vect{u},p\big)^{\trans}$.  
In the first problem (Riemann problem~1), taken from \cite{mignoneBodo_2005}, we use $\Gamma=4/3$ and
\begin{equation*}
  \vect{V}_{\leftState}=\big(1.0,0.9,0.0,0.0,1.0\big)^{\trans}\quad\text{and}\quad
  \vect{V}_{\rightState}=\big(1.0,0.0,0.0,0.0,10.0\big)^{\trans},
\end{equation*}
while in the second problem (Riemann problem~2), also from \cite{mignoneBodo_2005}, we use $\Gamma=5/3$, and let
\begin{equation*}
  \vect{V}_{\leftState}=\big(1.0,\vect{0.0},10^{3}\big)^{\trans}\quad\text{and}\quad
  \vect{V}_{\rightState}=\big(1.0,\vect{0.0},10^{-2}\big)^{\trans}.  
\end{equation*}
In both problems, the computational domain is $D=[0,1]$, the discontinuity is initially located at $x^{1}=0.5$, and the solutions are integrated to $t=0.4$.  
Results are plotted in Figure~\ref{fig:RelativisticHydro}.

\begin{figure}[h]
  \centering
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/MB2005_01_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/MB2005_04_Astronum_2018}
  \end{minipage} \\
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/MB2005_01_TCI_Astronum_2018}
  \end{minipage}\hspace{0.5pc}%
  \begin{minipage}{18pc}
    \includegraphics[width=18pc]{./Figures/MB2005_04_TCI_Astronum_2018}
  \end{minipage}
  \caption{\label{fig:RelativisticHydro}Results from solving the SR Euler equations with \thornado.  Riemann problem~1 (left panels) was solved using $100$ elements, while Riemann problem~2 (right panels) was solved using $400$ elements.  The numerical and exact solutions are plotted with plusses and solid lines, respectively.  In the lower panels we plot $xt$-plane locations of elements flagged for limiting}
\end{figure}

Again, the main features of the exact solution are captured with the DG method implemented in \thornado.  
For Riemann problem~1, we observe some oscillations in the density profile between the leftmost shock and the contact discontinuity located around $x^{1}=0.6$.  
For Riemann problem~2, the maximum density in the thin density shell between the shock and the contact discontinuity is significantly lower for the numerical solution than the exact solution.  
This is due to a combination of limited spatial resolution and the action of the slope limiter.  
We used $C_{\TCI}=0.03$ in both tests in this section, and we plot locations in the $xt$-plane of elements flagged for limiting in Figure~\ref{fig:RelativisticHydro}.  
The troubled-cell indicator tracks the discontinuities well for both Riemann problems.  
We have found the agreement with the exact solution to improve for larger values of $C_{\TCI}$, but at the expense of somewhat more oscillatory results.

@@

\section*{References}
\begin{thebibliography}{36}
  \bibitem{ascher_etal_1997} Ascher U M, Ruuth S J and Spiteri R J 1997 {\it Applied Numerical Mathematics} {\bf 23} 151
  \bibitem{bassi_etal_2013} Bassi F, Franchina N, Ghidoni A and Rebay S 2013 {\it Int. J. Numer. Meth. Fluids} {\bf 71} 1322
  \bibitem{blondin_etal_2003} Blondin J M, Mezzacappa A and DeMarino C 2003 {\it ApJ} {\bf 584} 980
  \bibitem{brennerScott_2008} Brenner S C and Scott L R 2008 {\it The Mathematical Theory of Finite Element Methods} Springer
  \bibitem{bruenn_etal_2018} Bruenn S W {\it et al} {\it Preprint} arXiv:1809.05608
  \bibitem{cockburnShu_1998} Cockburn B and Shu C-W 1998 {\it JCP} {\bf 141} 199
  \bibitem{cockburnShu_2001} Cockburn B and Shu C-W 2001 {\it Journal of Scientific Computing} {\bf 16} 173
  \bibitem{colellaWoodward_1984} Colella P and Woodward P 1984 {\it JCP} {\bf 54} 174
  \bibitem{dumbser_etal_2014} Dumbser M, Zanotti O, Loub{\`e}re R and Diot S 2014 {\it JCP} {\bf 278} 47
  \bibitem{fambri_etal_2018} Fambri F, Dumbser M, K{\"o}ppel S, Rezzolla L and Zanotti O 2018 {\it MNRAS} {\bf 477} 4543
  \bibitem{fuShu_2017} Fu G and Shu C-W 2017 {\it JCP} {\bf 347} 305
  \bibitem{harten_etal_1983} Harten A, Lax P D and Van Leer B 1983 {\it SIAM~Review} {\bf 25} 35
  \bibitem{hesthavenWarburton_2008} Hesthaven J S and Warburton T 2008 {\it Nodal discontinuous Galerkin methods: Algorithms, analysis and applications} Springer
  \bibitem{klockner_etal_2009} Kl{\"o}ckner A, Warburton T, Bridge J and Hesthaven J S 2009 {\it JCP} {\bf 228} 7863
  \bibitem{landauLifshitz_1979} Landau L D and Lifshitz E M 1979 {\it Fluid Mechanics} Pergamon
  \bibitem{liskaWendroff_2003} Liska R and Wendroff B 2003 {\it SIAM J. Sci. Comput.} {\bf 25} 3
  \bibitem{martiMuller_2003} Mart{\'i} J M and M{\"u}ller E 2003 {\it Living Rev. Relativ.} {\bf 6} 7
  \bibitem{mcnally_etal_2012} McNally C, Lyra W and Passy J-C 2012 {\it ApJS} {\bf 201} 18
  \bibitem{mignoneBodo_2005} Mignone A and Bodo G 2005 {\it MNRAS} {\bf 364} 126
  \bibitem{muller_2016} M{\"u}ller B 2016 {\it PASA} {\bf 33} 1
  \bibitem{qin_etal_2016} Qin T, Shu C-W and Yang Y 2016 {\it JCP} {\bf 315} 323
  \bibitem{radiceRezzolla_2011} Radice D and Rezzolla L 2011 {\it Phys. Rev. D} {\bf 84} 024010
  \bibitem{remacle_etal_2003} Remacle J-F, Flaherty J E and Shephard M S 2003 {\it SIAM~Review} {\bf 45} 53
  \bibitem{rezzollaZanotti_2013} Rezzolla L and Zanotti O 2013 {\it Relativistic Hydrodynamics} Oxford
  \bibitem{schaal_etal_2015} Schaal K, Bauer A, Chandrashekar P R{\"u}diger P, Klingenberg C and Springel V 2015 {\it MNRAS} {\bf 453} 4278
  \bibitem{sedov_1959} Sedov L I 1959 {\it Similarity and Dimensional Methods in Mechanics} Academic Press
  \bibitem{shuOsher_1988} Shu C-W and Osher S 1988 {\it JCP} {\bf 77} 439
  \bibitem{shuOsher_1989} Shu C-W and Osher S 1989 {\it JCP} {\bf 83} 32
  \bibitem{shu_1997} Shu C-W 1997 {\it NASA/CR-97-206253 ICASE Report No. 97-65}
  \bibitem{shu_2016} Shu C-W 2016 {\it JCP} {\bf 316} 598
  \bibitem{sod_1978} Sod G A 1978 {\it JCP} {\bf 27} 1
  \bibitem{teukolsky_2016} Teukolsky S A 2016 {\it JCP} {\bf 312} 333
  \bibitem{toro_etal_1994} Toro E F, Spruce M and Speares W 1994 {\it Shock Waves} {\bf 4} 25
  \bibitem{toro_1999} Toro E F 1999 {\it NUMERICA: A Library of Source Codes for Teaching, Research and Applications}
  \bibitem{wilson_etal_1996} Wilson J R, Mathews G J and Marronetti P 1996 {\it Phys. Rev. D} {\bf 54} 1317
  \bibitem{wuTang_2015} Wu K and Tang H 2015 {\it JCP} {\bf 298} 539
  \bibitem{yahilLattimer_1982} Yahil A and Lattimer J M 1982 {\it Supernovae: A Survey of Current Research} Reidel
  \bibitem{yahil_1983} Yahil A 1983 {\it ApJ} {\bf 265} 1047
  \bibitem{zhangShu_2010} Zhang X and Shu C-W 2010 {\it JCP} {\bf 229} 8918
\end{thebibliography}
